---
title: 'Data Analysis in R'
author: "Niko Seino"
date: "2023-05-05"
output: 
  html_document:
    toc: true
    theme: "cosmo"
---

## About This Project

This project contains an analysis of survey data containing 37 variables 
related to the health and demographics of survey participants.

My goal was to explore and visualize the data, then determine the best 
demographic variables for predicting depression. 

To do this, I focused on four specific variables: 

*ADDEPEV2*:
Shows whether or not participants have been diagnosed with a depressive disorder.
In the dataset, this variable is numeric but will need to be treated as nominal.
The possible responses are: 1=Yes, 2=No, 7=Don't know/Not sure, 9=Refused

*POORHLTH*:
Shows how many days in the past 30 that poor physical or mental health kept 
participant from doing usual day-to-day activities. 

The values in this variable are integer values referring to days between 1-30, 
and '88' is used to indicate "None", or 0 days. Other values include 77, 
meaning "Don't know",and 99, meaning "refused". 

*EDUCA*
Shows participants' level of education.This is a numeric variable with the 
numbers representing categories of education level. There are 6 categories 
1-6 starting with 1 representing no formal education or only kindergarten, 
up to 6 representing 4 or more years of college. An additional category, 9, 
represents participants who did not give an answer.

*CHILDREN*
Shows the number of children in each participant's household.
This is a numeric variable representing the number of children. 
The value has a range of 1-87, and 88 represents "none" or 0, which 
was important to note when doing statistical calculations. 
'99' represents those who did not give an answer.


## Data Preparation
```{r}
# Clear memory
rm(list = ls())

# Load packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lm.beta))

# Set working directory and load dataset
#setwd("~/DTSC650")
brfss <- read_csv("BRFSS2015_650.csv", show_col_types = FALSE)
```


## Data Cleaning

1. Transforming variables

```{r warning=FALSE}
#Viewing the "Poor Health" data:
options(scipen = 999)
ggplot(brfss, aes(x = brfss$POORHLTH, na.rm = TRUE)) +
  geom_histogram(binwidth = 3)

# The graph shows that the data is varied between 1-30 with no real outliers.
#   I can also see that the majority of responses were 88, or 0 days. 
#   To better represent this and clean up the graph, I will update the data
#   to show the zeros and also eliminate the 77s and 99s.

brfss$POORHLTH <- replace(brfss$POORHLTH, brfss$POORHLTH == 88, 0)
brfss <- brfss %>%
  filter(POORHLTH <= 30, na.rm = TRUE)

# Updated graph:
ggplot(brfss, aes(x = POORHLTH)) +
  geom_histogram(binwidth = 2)
# Now I can clearly see the most responses are at 0, and there are no tails on
#   either end of the graph indicating no outliers. 
```

```{r}
# ADDEPEV2: Transforming into a binary variable with options "Yes" or "no"
# Eliminating the non responses ('7's and '9's)
# Converting variables to factors with levels

brfss <- brfss %>%
  filter(ADDEPEV2 == 1 | ADDEPEV2 == 2)
brfss$ADDEPEV2 <- factor(brfss$ADDEPEV2, levels = c(1, 2), labels = c("Yes", "No"))
```

```{r}
# Converting variables to factors and removing non responses
brfss <- brfss %>%
  filter(EDUCA > 0, EDUCA < 7)

# Adding labels to decrease confusion
brfss$EDUCA <- factor(brfss$EDUCA,
                         levels = c("1", "2", "3", "4", "5", "6"),
                         labels = c("None", "Elementary", "Some High School",
                                    "High School Graduate", "Some College",
                                    "College Graduate"))
```


2. Removing Outliers

```{r message=FALSE, warning=FALSE}
# Creating a graph to view the "Children" data:
ggplot(brfss, aes(x = brfss$CHILDREN, na.rm = TRUE)) +
  geom_histogram(binwidth = 3)
# The 88 and 99 values are making the data difficult to read, so I need to
#   transform the 88 to 0 and remove the 99's since they are not relevant.
# I also want to know what the max is so I can better fit the graph.
brfss$CHILDREN <- replace(brfss$CHILDREN, brfss$CHILDREN == 88, 0)
brfss <- brfss %>%
  filter(CHILDREN <= 87)
max(brfss$CHILDREN) 
#max number of children in a household is 77. This seems odd and a possible outlier.

# Creating a boxplot to view quantiles
ggplot(brfss, aes(x = CHILDREN)) +
  geom_boxplot()
# Here we can see several possible outliers.

childrenupper <- quantile(brfss$CHILDREN, .9985, na.rm = TRUE)
# This shows that 99.85% of the values are at 6 or below. 
# Values above 6 are more than 3 standard deviations from the mean, indicating 
#   they are outliers. 

# Update the dataset with outliers removed 
children_out <- which(brfss$CHILDREN > 6)
length(children_out) 
brfss <- brfss[-children_out, ]

```


## Data Visualization 

```{r}
# Visualizing poor health days

Q12a <- ggplot(brfss, aes(x = POORHLTH)) +
  geom_bar() + 
  labs(title = "Days in the Past 30 in which Poor Health 
       Prevented Usual Daily Activities")
Q12a
```

```{r}
# Poor health days by depression dx

ggplot(brfss, aes(x = POORHLTH)) +
  geom_bar() +
  facet_wrap(~ ADDEPEV2) +
  labs(title = "Days in the Past 30 in which Poor Health
       Affected Daily Activities of People 
       With and Without Depression")
```

Interpretation: 

We can see that there are peaks at both ends, 0 and 30, meaning most people
either were unaffected by poor health in the last 30 days, or were affected
every day. This pattern is similar in those who have been told they have 
depression, however in this population the proportion of 0 poor health days 
is much lower, and the proportion of 30 poor health days is higher. 

```{r}
# Simple bar graph to view counts of participants with depression dx
ggplot(brfss, aes(x = ADDEPEV2, fill = ADDEPEV2)) +
  geom_bar() +
  labs(title = "Number of People in Survey Who Have Had Depression",
       x = "Had Depression")
```

```{r}
# Depression in relation to education level
Q12b <- ggplot(brfss, aes(x = ADDEPEV2, fill = EDUCA)) +
  geom_bar(position = "dodge") + 
  labs(title = 'Education Level of People Who Have Had Depression')
Q12b
```

Interpretation: 

Among Respondents who reported they have NOT been diagnosed with depression,
there is a higher proportion of people who have completed 4 or more years of
college. The proportions of other education levels are similar.

```{r}
# Graphing education levels and number of children
Q12c <- ggplot(brfss) +
  geom_bar(aes(x = EDUCA, fill = EDUCA)) +
  labs(title = "Education Level of Survey Participants", x = "Education Level") +
  theme(axis.text.x = element_text(angle = 90))
Q12c

Q12d <- ggplot(brfss) +
  geom_bar(aes(x = CHILDREN)) +
  labs(title = "Number of Children in Households of Survey Participants") +
  scale_x_continuous(breaks = seq(0, 6, 1))
Q12d
```

```{r}
# Creating a table to show frequencies of education level per # of children
edu_kids <- table(brfss$EDUCA, brfss$CHILDREN)
edu_kids
```

```{r}
# Plotting a comparison of education level vs number of children
ggplot(brfss, aes(x = EDUCA, fill = EDUCA)) +
  geom_bar() +
  scale_fill_brewer("blues") +
  facet_grid(~ CHILDREN) +
  labs(title = "Education Level vs Number of Children", x = "Education Level") +
  theme(axis.text.x = element_text(angle = 90))
```

Interpretation:  

The majority of participants had 0 children in their households, and for all
numbers of children (0-6), counts increased as education level increased, with
"College Graduate" being the most common education level. The one exception 
was participants who had 6 children--"Some College" was the most frequent
education level and "High School Graduate" the second most frequent. 

## Data Exploration

Conducting descriptive statistical calculations to gain further insights and 
answer some questions.

1. What is the average number of days poor health affected daily activities 
of people who completed only elementary school, vs those who graduated college?

```{r}
Q13a <- brfss %>%
  filter(EDUCA == "Elementary" | EDUCA == "College Graduate") %>%
  group_by(EDUCA) %>%
  summarize(mean_poorhlth = round(mean(POORHLTH), 2))
Q13a
```


2. What are the frequencies of each education level for participants who had
depression?

```{r}
Q13b <- brfss %>%
  filter(ADDEPEV2 == "Yes") %>%
  group_by(EDUCA) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
Q13b
```


3. What is the most common education level for participants who reported poor health
affected daily activities for more than 15 out of the past 30 days?

```{r}
Q13c <- brfss %>%
  filter(POORHLTH > 15) %>%
  group_by(EDUCA) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(1)
Q13c
```


4. What is the mean and standard deviation of number of children for those who
did vs did not report having depression?

```{r}
Q13d <- brfss %>%
  group_by(ADDEPEV2) %>%
  summarize(mean_children = mean(CHILDREN), sd_children = sd(CHILDREN)) %>%
  as.data.frame() %>%
  mutate_if(is.numeric, round, 2)
Q13d
```


## Using Data to Make Predictions

1. Creating a dataframe with just the variables I want to focus on

```{r}
df_lr <- brfss %>%
  select(POORHLTH, ADDEPEV2, EDUCA, CHILDREN)
```

2. Looking at summary stats

```{r}
summary(df_lr)
```


3. Performing logistic regressions to predict the likelihood of a participant 
having depression:

```{r}
# First, predicting depression from poor health days:
mod1 <- glm(ADDEPEV2 ~ POORHLTH, binomial(link = "logit"), df_lr)
summary(mod1)
# Converting coefficients from log odds to odds ratio
exp(coef(mod1))
```

Interpretation: 

POORHLTH odds ratio of .95 indicates that for each additional day of poor 
health, the odds of not having had depression decreases by about 5%. 
So there is a negative relationship between poor health days and not having 
had depression. The POORHLTH variable is statistically significant
in this model, however the AIC appears to be quite high. Will test other
models to see if we can get a lower AIC.

4. Adding another predictor variable

```{r}
mod2 <- glm(ADDEPEV2 ~ POORHLTH + EDUCA, binomial(), df_lr)
summary(mod2)
```


The EDUCA variable does not appear to be a good fit for this model as none of
the categories have statistically significant coefficients. 

5. Predicting depression from poor health days and number of children

```{r}
mod3 <- glm(ADDEPEV2 ~ POORHLTH + CHILDREN, binomial(), df_lr)
summary(mod3)
exp(coef(mod3))
```


The CHILDREN coefficient is statistically significant and has a negative
relationship with the outcome variable.

5. Choosing the model with the best predictor variables

```{r}
# Calculating AIC
AIC(mod3)
AIC(mod1)
```

Result: 

The AIC of our model with 2 predictors and the first model with just one predictor
are very close, but the AIC of the model with both POORHLTH and CHILDREN
is slightly lower, indicating it as the better model. 
